Biotic interactions structure zooplankton metacommunity dynamics following a summer heatwave

### 2024-02-29 writing/Manuscript.Rmd
updates based on edits from KD

---
title: "Biotic interactions structure zooplankton metacommunity dynamics following a summer heatwave"
author: "| Patrick L. Thompson$^1$^[Corresponding author: patrick.thompson@zoology.ubc.ca], Coreen Forbes$^1$^[forbes.coreen@gmail.com], Joey R. Bernhardt$^{1,2}$^[joey.bernhardt@uoguelph.ca], \n| Kaleigh E. Davis$^1$^[kdavis@zoology.ubc.ca], Keila A. Stark$^1$^[keilastark@zoology.ubc.ca], Evgeniya Yangel$^1$^[jane.s.yangel@gmail.com], Felipe E. Amadeo$^1$^[philamadeo@gmail.com],\n| Natalie E. Westwood$^1$^[westwood@zoology.ubc.ca], Mary I. O'Connor$^1$^[oconnor@zoology.ubc.ca] \n|\n|
  $^1$Biodiversity Research Center and the Department of Zoology,\n| University of British Columbia, Vancouver, B.C., Canada\n|
  $^2$Department of Integrative Biology,\n| University of Guelph, Guelph, Ontario, Canada\n"
date: "`r format(Sys.time(), '%B %d, %Y')`"
subtitle: "Running title: Heatwave metacommunity experiment"
output:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      keep_tex: false
csl: ecology-letters.csl
bibliography: TEAMM_Zoop.bib
always_allow_html: true
link-citations: yes
header-includes:
  \usepackage{gensymb}
  \usepackage[left]{lineno}
  \usepackage{setspace}\doublespacing
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}  
  \renewcommand{\thetable}{S\arabic{table}} 
  \setcounter{figure}{0} 
  \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

Biotic interactions are expected to cause community assembly and composition patterns to diverge from expectations based on abiotic conditions. The abundance and distribution of species depends on abiotic conditions as well as on the presence and abundance of other species that affect population sizes through competition, facilitation, or prediation [@hutchinson1957]. Competition and predation, for example, may prevent a population from establishing and growing when abiotic conditions would otherwise be suitable [@brown2014]. Thus, biotic interactions are a primary reason why communities that experience the same abiotic conditions may differ in composition [@davis1998; @ives2004; @thompson2021]. Despite this, most classic metacommunity theory [e.g., @loreau2003; @leibold2004] is expressed in simple terms (competition is the only species interaction) and employs the assumption that the strength of competition is equal amongst all individuals and species. Thus, much of this body of theory assumes that community composition is purely determined by abiotic conditions and dispersal. Metacommunity models that explicitly incorporate differences in strength of competition reveal the importance of biotic interactions in structuring communities, and make the prediction that biotic interactions reduce species sorting and lead to non-random patterns of species co-occurrence across space and time that are unexplained by local abiotic conditions [@thompson2020]. When communities experience periods of extreme abiotic conditions, such as during a heatwave, biotic interactions may prevent the community from returning to its initial compositional state [@davis1998; @thompson2021].

While the influence of dispersal and changing abiotic conditions, including temperature, on community composition in spatially structured landscapes can be easily assessed through experimental manipulation [e.g., @symons2013; @limberger2019], experimentally assessing the role of biotic interactions in such systems is more challenging. In microcosms, the role of biotic interactions in shaping community composition across abiotic gradients has been experimentally tested by comparing the response of species to abiotic change in monoculture vs. mixture [@davis1998; @thompson2021]. Because, in this experimental design, each species requires its own set of replicate experimental units [e.g., 216 microcosms in @thompson2021], scaling this approach up to larger scales such as mesocosms is prohibitive. Without being able to experimentally manipulate species interactions, an alternative is to control abiotic conditions and dispersal and then use statistical models to test whether community composition patterns can be explained by these the experimental treatments; if not, the alternative explanation is biotic interactions. Examples of such results would be positive or negative species co-occurrence patterns that are not explained by abiotic treatments as well as temporal shifts in community composition that are not explained by changes in abiotic conditions [@ovaskainen2020].

Here we present the results of a replicated pond zooplankton metacommunity experiment in which we tested how abiotic responses, dispersal, and biotic interactions shape the response and recovery of the community to a mid-summer heatwave. We test the following three hypotheses: H1 -- Community composition is primarily determined by abiotic conditions (here, temperature). If true, we predict that species abundances would vary strongly across spatial and temporal temperature gradients and community composition would be explained by temperature. H2 -- Dispersal modifies the effects of abiotic conditions on community composition. From this hypothesis, we predict that the abiotic environment explains variation in community composition when dispersal occurs at moderate rates, but not when dispersal rates are high. H3 -- Biotic interactions can lead to composition patterns not explained by species sorting. If true, we predict that community composition will vary among local communities such that there are negative and positive co-occurrence patterns over space and time that cannot be explained by abiotic conditions.

@@

Because our experimental design allowed us to control for abiotic differences other than those due to temperature, we infer that additional variation in species' abundances that is not associated with our fixed effects for temperature and dispersal is likely due to biotic interactions (but see discussion for caveats). We modelled variation that was unexplained by our treatments in three ways. First, we included time (sampling week) as a linear fixed effect in our model to account for temporal changes in species' abundances that were not associated with temperature of dispersal. We elected not to include time as a random or quadratic effect because the temperature was the highest in the middle of the three sampling weeks and so temperature and time would have been confounded in our model. Thus, we assumed that species' abundances that peaked or dipped in the middle sampling week compared to the first and last sample weeks were associated with temperature. Second, mesocosm identity was included as a random effect, allowing us to estimate whether species showed non-random patterns of co-occurrence within individual mesocosms that were not associated with temperature or dispersal. Finally, metacommunity identity was also included as a random effect, allowing us to estimate whether species showed non-random spatial patterns of co-occurrence at the metacommunity scale that were not associated with temperature or dispersal. Thus, if biotic interactions impede species sorting or result in successional dynamics (H3) then we expect to see strong time-abundance relationships and non-random species co-occurrence patterns at either the mesocosm or metacommunity scale.

@@

# Results
The community consisted of `r ncol(m$Y)` zooplankton taxa. The mean abundance per mesocosm across the three sampling weeks ranged from `r round(min(colMeans(m$Y))/10, 2)` individuals L$^{-1}$ for Ostracoda to `r round(max(colMeans(m$Y))/10, 1)` individuals L$^{-1}$ for Diaphanosoma, with a median value of `r round(median(colMeans(m$Y))/10, 1)` individuals L$^{-1}$ across all taxa.

@@

The accuracy of the HMSC model in predicting the species' abundances ranged from an R$^2$ of `r signif(min(MF$SR2), dig = 2)` for Chironomidae to `r signif(max(MF$SR2), dig = 2)` for Diaphanosoma, with a median value of `r signif(median(MF$SR2), dig = 2)`, across all taxa (Figure \@ref(fig:HMSCfig)a). Of this explained variation, an average of `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "temp * dispersal") %>% pull(variance) %>% round(3) *100`% was explained by the interaction between temperature and dispersal, `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "time") %>% pull(variance) %>% round(3) *100`% was explained by time, `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "mesocosm") %>% pull(variance) %>% round(3) *100`% was explained by mesocosm identity, and `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "metacommunity") %>% pull(variance) %>% round(3) *100`% was explained by metacommunity identity (Figure \@ref(fig:HMSCfig)a). Convergence of the HMSC model was acceptable with a mean scale reduction factor $\hat{R}$ of `r round(mean(converge.df$ge),3)` and with `r round(sum(converge.df$ge<1.1)/nrow(converge.df)*100)`% of parameters having values below 1.1.

Consistent with H1, all species showed strong and varied temperature--abundance relationships as is necessary for temperature driven species sorting (Figure \@ref(fig:tempcurves)). Taxa differed in the range of temperatures at which they were most abundant (Figure \@ref(fig:tempcurves)) as indicated by the varied varied linear temperature coefficients (Figure \@ref(fig:HMSCfig)b).  Daphnia and Calanoida had the coldest optima at 22.5 and 24.5 \degree C respectively and Chaoborus, Cyclopoida, and Diaphanosoma had the warmest optima, which were all above 30 \degree C. In addition, all taxa were estimated to have unimodal temperature--abundance relationships (Figure \@ref(fig:tempcurves)) as indicated by the negative quadratic temperature coefficients (Figure \@ref(fig:HMSCfig)b). The 95% credible intervals of these negative quadratic responses did not overlap zero, with the exception of Cyclopoida.

As expected under H2, dispersal did alter the realized effects of temperature on population abundance. The most common effect was that dispersal weakened the temperature--abundance relationships, as indicated by the positive dispersal by temperature$^2$ coefficients (Figure \@ref(fig:HMSCfig)b, Figure \@ref(fig:zoopgrid)). Contrary to H2, where we expected this to occur under high but not low dispersal, this weakening of the temperature--abundance relationships was equally common under both low and high dispersal, with five taxa showing 95% credible intervals that did not overlap zero in each dispersal rate. The less common effect of dispersal, exhibited by only two taxa under each dispersal rate, was a strengthening of the temperature-abundance relationship (i.e, negative dispersal by temperature$^2$ coefficients; Figure \@ref(fig:HMSCfig)b). Dispersal had more variable effects on the linear temperature responses, with both positive and negative dispersal temperature interactions being common at both dispersal rates (Figure \@ref(fig:HMSCfig)b). This indicates that dispersal was altering the temperatures at which abundances were highest, but not in a consistent way. Low dispersal caused Scapholeberis to  decreased and Chironomidae to increase in abundance but otherwise dispersal had limited direct influence on taxon abundances (Figure \@ref(fig:HMSCfig)b).

Consistent with H3, all species' abundances varied over time and there was evidence of non-random co-occurrence patterns that were not associated with the experimental treatments. Independent of the effects of temperature and dispersal, four taxa (Simocephalus, Daphnia, Chaoborus, Calanoida) were estimated to have decreased in abundance over time and the remaining six taxa were estimated to have increased in abundance over time (Figure \@ref(fig:HMSCfig)b). Within mesocosms, Cyclopoda were positively associated with Chironomidae and Simocephalus was negatively associated with Chironomidae and Cyclopoida (Figure \@ref(fig:HMSCfig)c). Within metacommunities, Diaphanosoma was negatively associated with Ostracoda, while Acari was positively associated with Cyclopoida and negatively associated with Scapholeberis (Figure \@ref(fig:HMSCfig)d).

The influence of biotic interactions emerges when comparing the composition of the post- and pre-heatwave communities (Figure \@ref(fig:nmds)). The post-heatwave communities tended to be comprised of Scapholeberis, Cyclopoda, Acari, Ostracoda, Chrimonidae, and Diaphanosoma; all species that tolerate high temperatures (Figure \@ref(fig:tempcurves)). These communities were not, however, as clearly structured by their temperature treatments as they were in week four, prior to the heatwave (Figure \@ref(fig:nmds)). Although temperatures were similar in the weeks preceding weeks 4 and 12, the composition of the communities differed considerably between the two time periods. Based on the variation not explained by temperature and dispersal in our model, we infer that these differences between weeks 4 and 12 are at least in part due to temporal changes in abundance that were independent of temperature and dispersal. This loss of temperature-driven differences in community composition following the heatwave and the temporal changes in species abundance that were not associated with temperature are counter to H1 and are consistent with H3. In addition, the non-random taxon co-occurrence patterns are evident in the post-heatwave communities which showed high compositional variability that was not associated with the temperature treatments, further supporting H3. Still, the influence of dispersal was evident in the week 12 communities. The composition of the no dispersal mesocosms was the most variable, the composition of mesocosms with low dispersal tended to remain dominated by Diaphanosoma, while the composition of the high dispersal treatments tended to have higher abundance of Scapholeberis, Ostracoda, Acari, and Cyclopoid but lower abundance of Calanoida and Daphnia.

@@

We found strong evidence that taxa differed in their temperature preferences, which resulted in spatial turnover in community composition across the temperature gradient within the metacommunities prior to the heatwave. These temperature responses also led to a predictable response of the community to the heatwave, whereby the communities tended to be dominated by a single heat tolerant taxa, Diaphanosoma. However, counter to expectations based on the species sorting paradigm [H1\; @leibold2004] the post-heatwave communities were not predictably structured based on the temperature preferences of taxa as they were prior to the heatwave, despite the fact that the temperature gradients within the metacommunities were maintained throughout the experiment. These post-heatwave communities tended to consist of taxa that could tolerate high temperatures, but the composition across replicate mesocosms was highly variable. We suggest that this variation is likely driven by interactions between taxa [following H3\; @davis1998; @thompson2020], which maintained and amplified compositional differences between communities, preventing the reestablishment of the temperature-based structure that characterized the metacommunities prior to the heatwave. In addition, counter to the spatial insurance hypothesis [H2\;@loreau2003; @gonzalez2009], low rates of dispersal did not facilitate species sorting. Instead, all rates of dispersal acted to reduce the degree to which temperature structured the metacommunities, resulting in compositional homogenization across space.

>>>>

---
title: "Biotic interactions structure zooplankton metacommunity dynamics following a summer heatwave"
author: "| Patrick L. Thompson$^1$^[Corresponding author: patrick.thompson@zoology.ubc.ca], Coreen Forbes$^1$^[forbes.coreen@gmail.com], Joey R. Bernhardt$^{1,2}$^[joey.bernhardt@uoguelph.ca], \n| Kaleigh E. Davis$^1$^[kdavis@zoology.ubc.ca], Keila A. Stark$^1$^[keilastark@zoology.ubc.ca], Evgeniya Yangel$^1$^[jane.s.yangel@gmail.com], Felipe E. Amadeo$^1$^[amadeo@zoology.ubc.ca],\n| Natalie E. Westwood$^1$^[westwood@zoology.ubc.ca], Mary I. O'Connor$^1$^[oconnor@zoology.ubc.ca] \n|\n|
  $^1$Biodiversity Research Center and the Department of Zoology,\n| University of British Columbia, Vancouver, B.C., Canada\n|
  $^2$Department of Integrative Biology,\n| University of Guelph, Guelph, Ontario, Canada\n"
date: "`r format(Sys.time(), '%B %d, %Y')`"
subtitle: "Running title: Heatwave metacommunity experiment"
output:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      keep_tex: false
csl: ecology-letters.csl
bibliography: TEAMM_Zoop.bib
always_allow_html: true
link-citations: yes
header-includes:
  \usepackage{gensymb}
  \usepackage[left]{lineno}
  \usepackage{setspace}\doublespacing
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}  
  \renewcommand{\thetable}{S\arabic{table}} 
  \setcounter{figure}{0} 
  \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

Biotic interactions are expected to cause community assembly and composition patterns to diverge from expectations based on abiotic conditions. The abundance and distribution of species depends on abiotic conditions as well as on the presence and abundance of other species that affect population sizes through competition, facilitation, or predation [@hutchinson1957]. Competition and predation, for example, may prevent a population from establishing and growing when abiotic conditions would otherwise be suitable [@brown2014]. Thus, biotic interactions are a primary reason why communities that experience the same abiotic conditions may differ in composition [@davis1998; @ives2004; @thompson2021]. Despite this, most classic metacommunity theory [e.g., @loreau2003; @leibold2004] is expressed in simple terms (competition is the only species interaction) and employs the assumption that the strength of competition is equal amongst all individuals and species. Thus, much of this body of theory assumes that community composition is purely determined by abiotic conditions and dispersal. Metacommunity models that explicitly incorporate differences in strength of competition reveal the importance of biotic interactions in structuring communities, and make the prediction that biotic interactions reduce species sorting and lead to non-random patterns of species co-occurrence across space and time that are unexplained by local abiotic conditions [@thompson2020]. When communities experience periods of extreme abiotic conditions, such as during a heatwave, biotic interactions may prevent the community from returning to its initial compositional state [@davis1998; @thompson2021].

While the influence of dispersal and changing abiotic conditions, including temperature, on community composition in spatially structured landscapes can be easily assessed through experimental manipulation [e.g., @symons2013; @limberger2019], experimentally assessing the role of biotic interactions in such systems is more challenging. In microcosms, the role of biotic interactions in shaping community composition across abiotic gradients has been experimentally tested by comparing the response of species to abiotic change in monoculture vs. polyculture [@davis1998; @thompson2021]. Because, in this experimental design, each species requires its own set of replicate experimental units [e.g., 216 microcosms in @thompson2021], scaling this approach up to larger scales such as mesocosms is prohibitive. Without being able to experimentally manipulate species interactions, an alternative is to control abiotic conditions and dispersal and then use statistical models to test whether community composition patterns can be explained by these the experimental treatments; if not, the alternative explanation is biotic interactions. Examples of such results would be positive or negative species co-occurrence patterns that are not explained by abiotic treatments as well as temporal shifts in community composition that are not explained by changes in abiotic conditions [@ovaskainen2020].

Here we present the results of a replicated pond zooplankton metacommunity experiment in which we tested how abiotic responses, dispersal, and biotic interactions shape the response and recovery of the community to a mid-summer heatwave. We test the following three hypotheses: H1 -- Community composition is primarily determined by abiotic conditions (here, temperature). If true, we predict that species abundances would vary strongly across spatial and temporal temperature gradients and community composition would be explained by temperature. H2 -- Dispersal modifies the effects of abiotic conditions on community composition. From this hypothesis, we predict that the abiotic environment explains variation in community composition when dispersal occurs at moderate rates, but not when dispersal rates are high. H3 -- Biotic interactions can lead to composition patterns not explained by species sorting. If true, we predict that community composition will vary among local communities such that there are negative and positive co-occurrence patterns over space and time that cannot be explained by abiotic conditions or dispersal.

@@

Because our experimental design allowed us to control for abiotic differences other than those due to temperature, we infer that additional variation in species' abundances that is not associated with our fixed effects for temperature and dispersal is likely due to biotic interactions (but see discussion for caveats). We modelled variation that was unexplained by our treatments in three ways. First, we included time (sampling week) as a linear fixed effect in our model to account for temporal changes in species' abundances that were not associated with temperature or dispersal. We elected not to include time as a random or quadratic effect because the temperature was the highest in the middle of the three sampling weeks and so temperature and time would have been confounded in our model. Thus, we assumed that species' abundances that peaked or dipped in the middle sampling week compared to the first and last sample weeks were associated with temperature. Second, mesocosm identity was included as a random effect, allowing us to estimate whether species showed non-random patterns of co-occurrence within individual mesocosms that were not associated with temperature or dispersal. Finally, metacommunity identity was also included as a random effect, allowing us to estimate whether species showed non-random spatial patterns of co-occurrence at the metacommunity scale that were not associated with temperature or dispersal. Thus, if biotic interactions impede species sorting or result in successional dynamics (H3) then we expect to see strong time-abundance relationships and non-random species co-occurrence patterns at either the mesocosm or metacommunity scale.

@@

# Results
The community consisted of `r ncol(m$Y)` zooplankton taxa. The mean abundance per mesocosm across the three sampling weeks ranged from `r round(min(colMeans(m$Y))/10, 2)` individuals L$^{-1}$ for Ostracoda to `r round(max(colMeans(m$Y))/10, 1)` individuals L$^{-1}$ for  *Diaphanosoma*, with a median value of `r round(median(colMeans(m$Y))/10, 1)` individuals L$^{-1}$ across all taxa.

@@

The accuracy of the HMSC model in predicting the species' abundances ranged from an R$^2$ of `r signif(min(MF$SR2), dig = 2)` for Chironomidae to `r signif(max(MF$SR2), dig = 2)` for  *Diaphanosoma*, with a median value of `r signif(median(MF$SR2), dig = 2)`, across all taxa (Figure \@ref(fig:HMSCfig)a). Of this explained variation, an average of `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "temp * dispersal") %>% pull(variance) %>% round(3) *100`% was explained by the interaction between temperature and dispersal, `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "time") %>% pull(variance) %>% round(3) *100`% was explained by time, `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "mesocosm") %>% pull(variance) %>% round(3) *100`% was explained by mesocosm identity, and `r VP.df %>% group_by(effect) %>% summarise(variance = mean(variance)) %>% filter(effect == "metacommunity") %>% pull(variance) %>% round(3) *100`% was explained by metacommunity identity (Figure \@ref(fig:HMSCfig)a). Convergence of the HMSC model was acceptable with a mean scale reduction factor $\hat{R}$ of `r round(mean(converge.df$ge),3)` and with `r round(sum(converge.df$ge<1.1)/nrow(converge.df)*100)`% of parameters having values below 1.1.

Consistent with H1, all species showed strong and varied temperature--abundance relationships as is necessary for temperature driven species sorting (Figure \@ref(fig:tempcurves)). Taxa differed in the range of temperatures at which they were most abundant (Figure \@ref(fig:tempcurves)) as indicated by the varied varied linear temperature coefficients (Figure \@ref(fig:HMSCfig)b).  *Daphnia* and Calanoida had the coldest optima at 22.5 and 24.5 \degree C respectively and  *Chaoborus*, Cyclopoida, and  *Diaphanosoma* had the warmest optima, which were all above 30 \degree C. In addition, all taxa were estimated to have unimodal temperature--abundance relationships (Figure \@ref(fig:tempcurves)) as indicated by the negative quadratic temperature coefficients (Figure \@ref(fig:HMSCfig)b). The 95% credible intervals of these negative quadratic responses did not overlap zero, with the exception of Cyclopoida.

As expected under H2, dispersal did alter the realized effects of temperature on population abundance. The most common effect was that dispersal weakened the temperature--abundance relationships, as indicated by the positive dispersal by temperature$^2$ coefficients (Figure \@ref(fig:HMSCfig)b, Figure \@ref(fig:zoopgrid)). Contrary to H2, where we expected this to occur under high but not low dispersal, this weakening of the temperature--abundance relationships was equally common under both low and high dispersal, with five taxa showing 95% credible intervals that did not overlap zero in each dispersal rate. The less common effect of dispersal, exhibited by only two taxa under each dispersal rate, was a strengthening of the temperature-abundance relationship (i.e, negative dispersal by temperature$^2$ coefficients; Figure \@ref(fig:HMSCfig)b). Dispersal had more variable effects on the linear temperature responses, with both positive and negative dispersal temperature interactions being common at both dispersal rates (Figure \@ref(fig:HMSCfig)b). This indicates that dispersal was altering the temperatures at which abundances were highest, but not in a consistent way. Low dispersal caused  *Scapholeberis* to  decreased and Chironomidae to increase in abundance but otherwise dispersal had limited direct influence on taxon abundances (Figure \@ref(fig:HMSCfig)b).

Consistent with H3, all species' abundances varied over time and there was evidence of non-random co-occurrence patterns that were not associated with the experimental treatments. Independent of the effects of temperature and dispersal, four taxa ( *Simocephalus*,  *Daphnia*,  *Chaoborus*, Calanoida) were estimated to have decreased in abundance over time and the remaining six taxa were estimated to have increased in abundance over time (Figure \@ref(fig:HMSCfig)b). Within mesocosms, Cyclopoda were positively associated with Chironomidae and  *Simocephalus* was negatively associated with Chironomidae and Cyclopoida (Figure \@ref(fig:HMSCfig)c). Within metacommunities,  *Diaphanosoma* was negatively associated with Ostracoda, while Acari was positively associated with Cyclopoida and negatively associated with  *Scapholeberis* (Figure \@ref(fig:HMSCfig)d).

The influence of biotic interactions emerges when comparing the composition of the post- and pre-heatwave communities (Figure \@ref(fig:nmds)). The post-heatwave communities tended to be comprised of  *Scapholeberis*, Cyclopoda, Acari, Ostracoda, Chironomidae, and  *Diaphanosoma*; all species that tolerate high temperatures (Figure \@ref(fig:tempcurves)). These communities were not, however, as clearly structured by their temperature treatments as they were in week four, prior to the heatwave (Figure \@ref(fig:nmds)). Although temperatures were similar in the weeks preceding weeks 4 and 12, the composition of the communities differed considerably between the two time periods. Based on the variation not explained by temperature and dispersal in our model, we infer that these differences between weeks 4 and 12 are at least in part due to temporal changes in abundance that were independent of temperature and dispersal. This loss of temperature-driven differences in community composition following the heatwave and the temporal changes in species abundance that were not associated with temperature are counter to H1 and are consistent with H3. In addition, the non-random taxon co-occurrence patterns are evident in the post-heatwave communities which showed high compositional variability that was not associated with the temperature treatments, further supporting H3. Still, the influence of dispersal was evident in the week 12 communities. The composition of the no dispersal mesocosms was the most variable, the composition of mesocosms with low dispersal tended to remain dominated by  *Diaphanosoma*, while the composition of the high dispersal treatments tended to have higher abundance of  *Scapholeberis*, Ostracoda, Acari, and Cyclopoid but lower abundance of Calanoida and  *Daphnia*.

@@

We found strong evidence that taxa differed in their temperature preferences, which resulted in spatial turnover in community composition across the temperature gradient within the metacommunities prior to the heatwave. These temperature responses also led to a predictable response of the community to the heatwave, whereby the communities tended to be dominated by a single heat tolerant taxa,  *Diaphanosoma*. However, counter to expectations based on the species sorting paradigm [H1\; @leibold2004] the post-heatwave communities were not predictably structured based on the temperature preferences of taxa as they were prior to the heatwave, despite the fact that the temperature gradients within the metacommunities were maintained throughout the experiment. These post-heatwave communities tended to consist of taxa that could tolerate high temperatures even though the cooler adapted taxa that were abundant prior to the heatwave were still present at low abundances. In addition, the composition across replicate mesocosms was highly variable. We suggest that this variation and the lack of temperature structured composition was driven by interactions between taxa [following H3\; @davis1998; @thompson2020]. These interactions likely maintained and amplified compositional differences between communities, preventing the reestablishment of the temperature-based structure that characterized the metacommunities prior to the heatwave. In addition, counter to the spatial insurance hypothesis [H2\;@loreau2003; @gonzalez2009], low rates of dispersal did not facilitate species sorting. Instead, all rates of dispersal acted to reduce the degree to which temperature structured the metacommunities, resulting in compositional homogenization across space.
